Evidence for a Color Parameter Unassociated With Dust Within the Type~Ia Supernovae of the Nearby Supernova Factory

### 2017-10-27 submit/apj.tex
changes for Manu

@@

The parameterized intrinsic dispersion, $C_c$, seemingly introduces degeneracy in the model, as magnitude and color variation
ascribed to $\Delta$, $\vec{\gamma}^0 k_0$, and $\vec{\gamma}^1 k_1$ could also be attributed to intrinsic dispersion.  There are several features of the model
that drive the assignation of variations away from $C_c$:  Maximizing the posterior disfavors the increase of $\det{(C_c)}$
\color{red}
and hence favors solutions of $\gamma$'s that minimize the residual dispersion described by $C_c$;
\color{black}
The distributions of $\Delta$, $k_0$, and $k_1$ turn out to
be non-Gaussian, and so are not well described by the Normal covariance we impose in $C_c$; the grey magnitude offsets, $\Delta$, would appear as a constant
in all elements of the covariance matrix, which is disfavored for the Bayesian prior selected for $C_c$.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=5.2in]{residual.pdf} 
            \caption{
            \color{red}
            Model-expectation colors (including  supernova-specific spectral features and $\gamma k$ terms)
            versus observed colors, for four selected colors. 
            \color{black}
            Plotted on the right for each color
is the histogram of the difference between the expected and
observed
 colors.            \color{red}
(Note that the $\Delta$ in the plot signifies the difference in colors, not the per-supernova magnitude offset.)
            \color{black}
            \label{residual:fig}}
\end{figure}

@@

Non-trivial residual magnitude
and color dispersions are captured in $C_c$
at the level of 4--10$\sigma$.
The square
root of the diagonal elements are denoted as $\sigma$, which
range from
$\sim 0.01$ to 0.06 mag, significantly smaller
than the dispersion in $\Delta$.
Of particular interest within $C_c$ is intrinsic color dispersion.  The intrinsic dispersions in ${\hat{V}}$ and colors with respect to ${\hat{V}}$ are
\begin{multline}
\sigma({\hat{V}},{\hat{U}}-{\hat{V}},{\hat{B}}-{\hat{V}},{\hat{V}}-{\hat{R}},{\hat{V}}-{\hat{I}})=\\
\left(
0.021^{+0.004}_{-0.005} ,\, 
0.067^{+0.010}_{-0.010},\, 
0.034^{+0.005}_{-0.005},\, 
0.023^{+0.002}_{-0.002},\, 
0.053^{+0.004}_{-0.004}
\right)
 \text{mag}.
 \label{sig_intrinsic:eqn}
 \end{multline}
The correlations between these parameters are
\begin{multline}
Cor({\hat{V}},{\hat{U}}-{\hat{V}},{\hat{B}}-{\hat{V}},{\hat{V}}-{\hat{R}},{\hat{V}}-{\hat{I}})=\\
\begin{pmatrix}
\begin{array}{rrrrr}
\hat{V} & -0.48^{+0.20}_{-0.18} & -0.34^{+0.21}_{-0.22} & 0.88^{+0.10}_{-0.25} & 0.60^{+0.11}_{-0.19} \\
\ldots & {\hat{U}}-{\hat{V}} & 0.60^{+0.10}_{-0.13} & -0.38^{+0.18}_{-0.17} & -0.33^{+0.15}_{-0.14} \\
\ldots  & \ldots & {\hat{B}}-{\hat{V}} & -0.21^{+0.17}_{-0.17} & 0.08^{+0.14}_{-0.15} \\
\ldots  & \ldots  &\ldots  & {\hat{V}}-{\hat{R}} & 0.65^{+0.06}_{-0.07} \\
\ldots  & \ldots  &\ldots  &\ldots & {\hat{V}}-{\hat{I}} \\
\end{array}
 \end{pmatrix}.
 \label{cor_intrinsic:eqn}
 \end{multline}
To reduce clutter only the upper triangular part of the otherwise symmetric matrix is shown.
These color variances are smaller than those derived by \citet{2003A&A...404..901N} or \citet{2007ApJ...659..122J}, for cases where
a direct comparison can be made.
\citetalias{2017ApJ...842...93M}  find an intrinsic dispersion of the SALT2 $c$ parameter (associated with ${\hat{B}}-{\hat{V}}$) of $\approx 0.07$ mag,
though variation of dust properties accounts for some of that dispersion.
%model11_rubin.py
Having no intrinsic color dispersion is disfavored, and not allowing for them can have important consequences.  Running an analysis assuming no intrinsic dispersion (replacing the probabilistic intrinsic magnitudes in Eq.~\ref{ewsiv:eqn}
with a strict equality), we find significant shifts in all slope parameters  and 
artificially
smaller statistical uncertainties.
\color{red}
Note that underestimated measurement uncertainties in $C$ could contribute to some of the dispersion in what we have considered as an intrinsic
$C_c$; we do not explore this source of systematic error in this article.
\color{black}

@@

The resulting credible intervals of our parameters are given in Table~\ref{global2:tab}.  Comparison with the
results of the 2-parameter extrinsic model given in Table~\ref{global:tab} shows that  parameters that are common to both
are consistent  to well within 1$\sigma$ uncertainties.  The most significant change is in the reduction of the standard
deviations of the magnitude residuals $\sigma_X$
in the ${\hat{U}}$ and ${\hat{B}}$ bands.

@@

Potential biases in our results are measured through simulation.
We generate 100 random realizations of data with input  parameters equal to the median values found in
this Section, except for $\vec{\delta}$, which is assigned
the mode shown in Figure~\ref{deltacorner:fig},
and the same noise as our data.  The posterior for each realization is analyzed individually, and the posteriors of all realizations are stacked.  The stack is sensitive to
biases in the best fit value.
The stack of  $\vec{\delta}$  is shown in Figure~\ref{simdelta:fig}.  There is a bias between
the input  $(-1.12, -0.89, -0.36,  -0.16, 0.21)$ and the mode of  the stacks $(-1.2, -1.06, -0.49, -0.25, 0.21)$.
We do not consider this bias to be an issue in our detection of a third parameter, as the difference is significantly
smaller than the statistical uncertainty of a single analysis and the length of the vector changes by $< 3\%$.
The posterior stacks for the other global parameters 
are shown in Figures~\ref{simglobal1:fig}--\ref{simglobal5:fig}.  The input parameters are recovered with negligible bias
relative to the statistical uncertainties. 
\color{red}
The standard deviations of $C_c$, $\vec{\sigma}$,  are not biased high, which would have been a signature of leakage of
intrinsic color variations from $\delta$ into intrinsic dispersion.
\color{black}
Note that some bias is expected: the priors for the distributions of $\Delta$, $k$, and $D$ are inconsistent
with the extracted distributions.  Also, 
there is a tendency for the slope parameters associated with the spectral features to be slightly underestimated, which is an expected consequence of 
the uncertainties in the predictor variable that lead to 
regression dilution \citep{spearman04}.

>>>>

\label{modelmodel:sec}

@@

The parameterized intrinsic dispersion, $C_c$,  introduces degeneracy in the model, as magnitude and color variation
ascribed to $\Delta$, $\vec{\gamma}^0 k_0$, and $\vec{\gamma}^1 k_1$ can  be attributed to intrinsic dispersion
\color{blue}
and vice versa.
In the case where there is independent intrinsic dispersion in all 5 bands, i.e.\  the input $C_c$ has rank-5, there exists degenerate solutions with a $C_c$ of rank 2.  Those identical predictions generally do not yield the same likelihood, as the determinant of $C_c$ varies over those solutions.  In a maximum-likelihood analysis the rank-2 is preferred over the rank-5 solution, resulting in some intrinsic dispersion getting attributed to $\Delta$, $k_0$, and $k_1$.
In our Bayesian analysis we impose priors on $C_c$, so that a zero-value determinant is disfavored.    
Thus, leakage from per-supernova intrinsic color parameters to intrinsic dispersion could bias the intrinsic color vectors of the SN Ia family,  or conversely, intrinsic dispersion could be spuriously attributed to an intrinsic color.  We examine this effect in the discussion of the results in \S\ref{model2:sec}.
\color{black}
%
%There are several features of the model
%that drive the assignation of variations away from $C_c$:  Maximizing the posterior disfavors the increase of $\det{(C_c)}$
%\color{red}
%and hence favors solutions of $\gamma$'s that minimize the residual dispersion described by $C_c$;
%\color{black}
%The distributions of $\Delta$, $k_0$, and $k_1$ turn out to
%be non-Gaussian, and so are not well described by the Normal covariance we impose in $C_c$; the grey magnitude offsets, $\Delta$, would appear as a constant
%in all elements of the covariance matrix, which is disfavored for the Bayesian prior selected for $C_c$.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=5.2in]{residual.pdf} 
            \caption{
            \color{red}
            Model-expectation colors (including  supernova-specific spectral features and $\gamma k$ terms)
            versus observed colors, for four selected colors. 
            \color{black}
            Plotted on the right for each color
is the histogram of the difference between the expected and
observed
 colors. 
            \label{residual:fig}}
\end{figure}

@@

Non-trivial residual magnitude
and color dispersions are captured in $C_c$
at the level of 4--10$\sigma$.
The square
root of the diagonal elements are denoted as $\sigma$, which
range from
$\sim 0.01$ to 0.06 mag, significantly smaller
than the dispersion in $\Delta$.
Of particular interest within $C_c$ is intrinsic color dispersion.  The intrinsic dispersions in ${\hat{V}}$ and colors with respect to ${\hat{V}}$ are
\begin{multline}
\sigma({\hat{V}},{\hat{U}}-{\hat{V}},{\hat{B}}-{\hat{V}},{\hat{V}}-{\hat{R}},{\hat{V}}-{\hat{I}})=\\
\left(
0.021^{+0.004}_{-0.005} ,\, 
0.067^{+0.010}_{-0.010},\, 
0.034^{+0.005}_{-0.005},\, 
0.023^{+0.002}_{-0.002},\, 
0.053^{+0.004}_{-0.004}
\right)
 \text{mag}.
 \label{sig_intrinsic:eqn}
 \end{multline}
The correlations between these parameters are
\begin{multline}
Cor({\hat{V}},{\hat{U}}-{\hat{V}},{\hat{B}}-{\hat{V}},{\hat{V}}-{\hat{R}},{\hat{V}}-{\hat{I}})=\\
\begin{pmatrix}
\begin{array}{rrrrr}
\hat{V} & -0.48^{+0.20}_{-0.18} & -0.34^{+0.21}_{-0.22} & 0.88^{+0.10}_{-0.25} & 0.60^{+0.11}_{-0.19} \\
\ldots & {\hat{U}}-{\hat{V}} & 0.60^{+0.10}_{-0.13} & -0.38^{+0.18}_{-0.17} & -0.33^{+0.15}_{-0.14} \\
\ldots  & \ldots & {\hat{B}}-{\hat{V}} & -0.21^{+0.17}_{-0.17} & 0.08^{+0.14}_{-0.15} \\
\ldots  & \ldots  &\ldots  & {\hat{V}}-{\hat{R}} & 0.65^{+0.06}_{-0.07} \\
\ldots  & \ldots  &\ldots  &\ldots & {\hat{V}}-{\hat{I}} \\
\end{array}
 \end{pmatrix}.
 \label{cor_intrinsic:eqn}
 \end{multline}
To reduce clutter only the upper triangular part of the otherwise symmetric matrix is shown.
These color variances are smaller than those derived by \citet{2003A&A...404..901N} or \citet{2007ApJ...659..122J}, for cases where
a direct comparison can be made.
\citetalias{2017ApJ...842...93M}  find an intrinsic dispersion of the SALT2 $c$ parameter (associated with ${\hat{B}}-{\hat{V}}$) of $\approx 0.07$ mag,
though variation of dust properties accounts for some of that dispersion.
%model11_rubin.py
Having no intrinsic color dispersion is disfavored, and not allowing for them can have important consequences.  Running an analysis assuming no intrinsic dispersion (replacing the probabilistic intrinsic magnitudes in Eq.~\ref{ewsiv:eqn}
with a strict equality), we find significant shifts in all slope parameters  and 
artificially
smaller statistical uncertainties.

@@

The resulting credible intervals of our parameters are given in Table~\ref{global2:tab}.  
\color{blue}
The $\delta$ parameters are normalized to $\hat{U}$ instead of $\hat{V}$, since the posterior of $\delta_{\hat{V}}$ covers zero.
\color{black}
Comparison with the
results of the 2-parameter extrinsic model given in Table~\ref{global:tab} shows that  parameters that are common to both
are consistent  to well within 1$\sigma$ uncertainties.  The most significant change is in the reduction of the standard
deviations of the magnitude residuals $\sigma_X$
in the ${\hat{U}}$ and ${\hat{B}}$ bands.
\color{blue}
The variance captured by the first two eigenvectors of $C_c$ from Model~II is $4.7 \times 10^{-3}$ mag$^2$, as
compared to  $6.4 \times 10^{-3}$ mag$^2$ from Model~I; this reduction shows the amount of intrinsic dispersion in Model~I
that  is now attributed to the explicit $\delta D$ term.
\color{black}

@@

Potential biases in our results are measured through simulation.
We generate 100 random realizations of data with input  parameters equal to the median values found in
this Section, except for $\vec{\delta}$, which is assigned
the mode shown in Figure~\ref{deltacorner:fig},
and the same noise as our data.  The posterior for each realization is analyzed individually, and the posteriors of all realizations are stacked.  The stack is sensitive to
biases in the best fit value.
The stack of  $\vec{\delta}$  is shown in Figure~\ref{simdelta:fig}.  There is a bias between
the input  $(-1.12, -0.89, -0.36,  -0.16, 0.21)$ and the mode of  the stacks $(-1.2, -1.06, -0.49, -0.25, 0.21)$.
We do not consider this bias to be an issue in our detection of a third parameter, as the difference is significantly
smaller than the statistical uncertainty of a single analysis and the length of the vector changes by $< 3\%$.
The posterior stacks for the other global parameters 
are shown in Figures~\ref{simglobal1:fig}--\ref{simglobal5:fig}.  The input parameters are recovered with negligible bias
relative to the statistical uncertainties. 
Note that some bias is expected: the priors for the distributions of $\Delta$, $k$, and $D$ are inconsistent
with the extracted distributions.  Also, 
there is a tendency for the slope parameters associated with the spectral features to be slightly underestimated, which is an expected consequence of 
the uncertainties in the predictor variable that lead to 
regression dilution \citep{spearman04}.

\color{blue}
We now return to a degeneracy of our models mentioned in \S\ref{modelmodel:sec}.
Parameters are susceptible to the misassignment, in either direction, between intrinsic color, $\Delta$, $k$, and $D$, and intrinsic dispersion $C_c$.
The first two eigenvectors of account for $0.89_{-0.05}^{+0.04}$ of the variance of $C_c$, meaning
that potential leakage of intrinsic colors can account for less than 10\% of the variance, corresponding to 0.024 mag standard deviation.
We establish through simulations that the standard deviations of $C_c$, $\vec{\sigma}$,  are not biased high, which would have been a signature of leakage of
intrinsic color variations from $\delta$ into intrinsic dispersion.
The simulations of input null intrinsic color terms and input intrinsic dispersion do not produce spurious detections of intrinsic color vectors.  We thus conclude that
biases induced by this degeneracy are low relative to our statistical uncertainties.
\color{black}
